Algebraic Principles for Concurrency Verification Tools

### 2013-10-06 Paper/paper.tex
Added intuitive explanations of rules

\documentclass{llncs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{graphicx,color}
\usepackage{multicol}
\usepackage{bussproofs}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}

@@

\begin{abstract}
We investigate the laws required for the derivation of the
rely/guarantee and owicki-gries inference rules in a minimalistic
algebraic setting based on Kleene algebra...
\end{abstract}

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebra for Rely/Guarantee}
\label{sec:RG}

Define a rely/guarantee algebra as a structure
$(K,RG,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
lattice, $(K,+,\cdot,\|,0,1)$ is a weak trioid and
$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. $RG$ is a distinguished
subset of relys/guarantees which satisfy the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|xy &= (r\|x)(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $RG$, and
$x,y,z$ for arbitrary elements of $K$. Some elementary consequences of these rules are as follows
\begin{align*}
1 &\le r,\\
r^\star = rr &= r = r\|r,\\
r\|x^+ &= (r\|x)^+.
\end{align*}

Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
show this, we use Isabelle's
\emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample generator
to construct models which violate each of these axioms, yet satisfy
all others. The models thus constructed for (\ref{rg1}) and
(\ref{rg3}) are shown in Figures \ref{fig:rg1} and \ref{fig:rg3}
respectively. The model constructed for (\ref{rg2}) is just the two
element rely/guarantee algebra with $0$ and $1$.

\begin{figure}[H]
\centering
\begin{subfigure}{0.24\textwidth}
\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
  \node (center) {};
  \node [elem] (r1) at (90:1) {$r_1$};
  \node [elem] (one) at (200:1) {$1$};
  \node [elem] (r2) at (20:1) {$r_2$};
  \node [elem] (zero) at (-90:1) {$0$};

  \path [line] (r1) -- (r2);
  \path [line] (r1) -- (one);
  \path [line] (r1) -- (zero);
  \path [line] (r2) -- (one);
  \path [line] (r2) -- (zero);
  \path [line] (one) -- (zero);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}{0.24\textwidth}
\begin{align*}
r_1 \| r_1 &= r_1\\
r_1 \| r_2 &= r_1\\
r_2 \| r_1 &= r_1\\
r_2 \| r_2 &= r_1
\end{align*}
\end{subfigure}
\begin{subfigure}{0.24\textwidth}
\begin{align*}
r_1r_1 &= r_1\\
r_1r_2 &= r_1\\
r_2r_1 &= r_1\\
r_2r_2 &= r_2
\end{align*}
\end{subfigure}
\begin{subfigure}{0.24\textwidth}
\begin{align*}
r_1^\star &= r_1\\
r_2^\star &= r_2
\end{align*}
\end{subfigure}
\caption{4 element counterexample for $r \in RG \implies r\|r \le r$}
\label{fig:rg1}
\end{figure}

@@

\begin{figure}
\centering
\begin{subfigure}{0.24\textwidth}
\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
  \node (center) {};
  \node [elem] (r1) at (90:1) {$r_1$};
  \node [elem] (one) at (0:1) {$1$};
  \node [elem] (zero) at (-90:1) {$0$};

  \path [line] (r1) -- (one);
  \path [line] (r1) -- (zero);
  \path [line] (r1) -- (zero);
  \path [line] (one) -- (zero);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}{0.24\textwidth}
\begin{align*}
r_1 \| r_1 &= r_1\\
r_1 r_1 &= r_1\\
r_1^\star &= r_1
\end{align*}
\end{subfigure}
\caption{3 element counterexample for $r \in RG \implies r\|xy = (r\|x)(r\|y)$}
\label{fig:rg3}
\end{figure}

Jones quintuples can be encodeded in this setting as
\begin{align}
r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g.
\end{align}
Using this encoding we can derive all the standard rules for
rely/guarantee, shown in Figure \ref{fig:rgrules}.

We can expand our rely/guarantee algebras to have residuals
axiomatised by Galois connections for both parallel and sequential
composition, i.e.
\begin{align}
xy \le z &\iff y \le x \rightarrow z, \label{res1}\\
xy \le z &\iff x \le z \leftarrow y, \label{res2}\\
x\|y \le z &\iff y \le x/z \label{res3}.
\end{align}
For sequential composition, (\ref{res1}) and (\ref{res2}) essentially
provide a link between weakest-precondition style semantics and hoare
logic. As mentioned in Section \ref{sec:KA}, the triple $\{p\} c
\{q\}$ can be encoded as $pc \le q$, while $q \leftarrow c$ represents
$wlp(c,q)$. Equation (\ref{res2}) can thus be written as $\{p\} c
\{q\} \iff p \le wlp(c,q)$. The parallel residual $x/z$ can be
understood similarly as follows; it is the weakest program such that
when placed in parallel with $x$, the composition behaves as $z$.

@@

\begin{prooftree}
\RightLabel{Star}
\AxiomC{$pr \le p$}
\AxiomC{$r, g \vdash \{p\}x\{p\}$}
\BinaryInfC{$r, g \vdash \{p\}x^\star\{p\}$}
\end{prooftree}
\caption{Rely/guarantee inference rules}
\label{fig:rgrules}
\end{figure}
x

>>>>

\documentclass{llncs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{graphicx,color}
\usepackage{multicol}
\usepackage{bussproofs}
\usepackage{float}

@@

\begin{abstract}
We investigate the laws required for the derivation of the
rely-guarantee and owicki-gries inference rules in a minimalistic
algebraic setting based on Kleene algebra...
\end{abstract}

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebra for Rely-Guarantee}
\label{sec:RG}

Define a rely-guarantee algebra as a structure
$(K,RG,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
lattice, $(K,+,\cdot,\|,0,1)$ is a weak trioid and
$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. $RG$ is a distinguished
subset of relys/guarantees which satisfy the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|xy &= (r\|x)(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $RG$, and
$x,y,z$ for arbitrary elements of $K$. Some elementary consequences of these rules are as follows
\begin{align*}
1 &\le r,\\
r^\star = rr &= r = r\|r,\\
r\|x^+ &= (r\|x)^+.
\end{align*}

%% Could use better explanations.

Axiom (\ref{rg1}) states that interference from the a rely condition
being run twice in parallel, is no different from just the
interference from that rely condition. Axiom (\ref{rg2}) states that
interference from a single rely condition is less than interference
from itself and another rely condition. Axiom (\ref{rg3}) allows a
rely or guarantee to be split across sequential programs. Axiom
(\ref{rg4}) is similar to Axiom (\ref{rg3}) in intent, except it deals
with finite iteration.

Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
show this, we use Isabelle's
\emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample generator
to construct models which violate each of these axioms, yet satisfy
all others. The models thus constructed for (\ref{rg1}) and
(\ref{rg3}) are shown in Figures \ref{fig:rg1} and \ref{fig:rg3}
respectively. The model constructed for (\ref{rg2}) is just the two
element rely-guarantee algebra with $0$ and $1$.

\begin{figure}[H]
\centering
\begin{minipage}{0.24\textwidth}
\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
  \node (center) {};
  \node [elem] (r1) at (90:1) {$r_1$};
  \node [elem] (one) at (200:1) {$1$};
  \node [elem] (r2) at (20:1) {$r_2$};
  \node [elem] (zero) at (-90:1) {$0$};

  \path [line] (r1) -- (r2);
  \path [line] (r1) -- (one);
  \path [line] (r1) -- (zero);
  \path [line] (r2) -- (one);
  \path [line] (r2) -- (zero);
  \path [line] (one) -- (zero);
\end{tikzpicture}
\end{minipage}
\begin{minipage}{0.24\textwidth}
\begin{align*}
r_1 \| r_1 &= r_1\\
r_1 \| r_2 &= r_1\\
r_2 \| r_1 &= r_1\\
r_2 \| r_2 &= r_1
\end{align*}
\end{minipage}
\begin{minipage}{0.24\textwidth}
\begin{align*}
r_1r_1 &= r_1\\
r_1r_2 &= r_1\\
r_2r_1 &= r_1\\
r_2r_2 &= r_2
\end{align*}
\end{minipage}
\begin{minipage}{0.24\textwidth}
\begin{align*}
r_1^\star &= r_1\\
r_2^\star &= r_2
\end{align*}
\end{minipage}
\caption{4 element counterexample for $r \in RG \implies r\|r \le r$}
\label{fig:rg1}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{minipage}{0.24\textwidth}
\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
  \node (center) {};
  \node [elem] (r1) at (90:1) {$r_1$};
  \node [elem] (one) at (0:1) {$1$};
  \node [elem] (zero) at (-90:1) {$0$};

  \path [line] (r1) -- (one);
  \path [line] (r1) -- (zero);
  \path [line] (r1) -- (zero);
  \path [line] (one) -- (zero);
\end{tikzpicture}
\end{minipage}
\begin{minipage}{0.24\textwidth}
\begin{align*}
r_1 \| r_1 &= r_1\\
r_1 r_1 &= r_1\\
r_1^\star &= r_1
\end{align*}
\end{minipage}
\caption{3 element counterexample for $r \in RG \implies r\|xy = (r\|x)(r\|y)$}
\label{fig:rg3}
\end{figure}

Jones quintuples can be encodeded in this setting as
\begin{align}
r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g.
\end{align}
Using this encoding we can derive all the standard rules for
rely-guarantee, shown in Figure \ref{fig:rgrules}.

We can expand our rely-guarantee algebras to have residuals
axiomatised by Galois connections for both parallel and sequential
composition, i.e.
\begin{align}
xy \le z &\iff y \le x \rightarrow z, \label{res1}\\
xy \le z &\iff x \le z \leftarrow y, \label{res2}\\
x\|y \le z &\iff y \le x/z \label{res3}.
\end{align}
For sequential composition, (\ref{res1}) and (\ref{res2}) essentially
provide a link between weakest-precondition style semantics and hoare
logic. As mentioned in Section \ref{sec:KA}, the triple $\{p\} c
\{q\}$ can be encoded as $pc \le q$, while $q \leftarrow c$ represents
$wlp(c,q)$. Equation (\ref{res2}) can thus be written as $\{p\} c
\{q\} \iff p \le wlp(c,q)$. The parallel residual $x/z$ can be
understood similarly as follows; it is the weakest program such that
when placed in parallel with $x$, the composition behaves as $z$.

@@

\begin{prooftree}
\RightLabel{Star}
\AxiomC{$pr \le p$}
\AxiomC{$r, g \vdash \{p\}x\{p\}$}
\BinaryInfC{$r, g \vdash \{p\}x^\star\{p\}$}
\end{prooftree}
\caption{Rely-guarantee inference rules}
\label{fig:rgrules}
\end{figure}
x
